---
title: "Social Data Science" 
subtitle: "Data structures"
author: Dr. Thomas Davidson
institute: Rutgers University
date: September 7, 2021
output:
    beamer_presentation:
      theme: "Szeged"
      colortheme: "beaver"
      fonttheme: "structurebold"
      toc: false
      incremental: false
header-includes:
  - \usepackage{multicol}
  - \usepackage{hyperref}
  - \usepackage{caption}
  - \captionsetup[figure]{font=scriptsize}
  - \captionsetup[figure]{labelformat=empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(dev = 'pdf')
library("knitr")
library("formatR")

opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
opts_chunk$set(tidy = FALSE)

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```


# Plan
- Object-oriented programming
- Basic types
- Vectors
- Lists
- Matrices
- Data frames and tibbles
- A note on style

# Object-oriented programming
- A paradigm of computer programming
    - We create *objects* of different *classes* such as numbers, strings, and data frames
    - These objects have *attributes*, properties such as data
        - e.g. The numeric object we call **`A`** has an attribute called `value` equal to **`1`**
    - Objects are associated with *methods* that allow us to manipulate them
        - e.g. a numeric object might have a method called `add`, such that **`A + A`** will return **`2`**.

# Basic types
There are four basic types we will be using throughout the class. We use the `<-` operator to assign an object to a name.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
# Character (also known as "strings")
name <- "Gary"
# Numeric ("float" in Python)
weight <- 13.2
# Integer ("int" for short)
age <- 4L
# Logical
human <- FALSE
```
\tiny The other two are called `complex` and `raw`. See documentation: https://cran.r-project.org/doc/manuals/R-lang.html

# Basic types
There are a few useful commands for inspecting objects.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(name) # Prints value in console
class(name) # Shows class of object
typeof(name) # Shows type of object, not always equal to class
```

# Basic types
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(weight) # Prints value in console
class(weight) # Shows class of object
typeof(weight) # Shows type of object, not always equal to class
```

# Basic types
We can also use the `==` expression to verify the value of an object. We will discuss this in more detail next lecture.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
name == "Gary" 
age == 3L
age >= 3L # is greater than
age != 3L # is not
```

# Vectors
A vector is a collection of elements of the *same* class
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
# We can define an empty vector with N elements of a class
x <- logical(5)
print(x)
y <- numeric(5)
print(y)
z <- character(5)
print(z)
```

# Vectors
Let's take a closer look at numeric vectors. We can use the "combine" function `c()` to concatenate values into a vectr.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1 <- c(1,2,3,4,5)
v2 <- c(1,1,1,1,1)
class(v1) # check the class of this vector
v1 + v2 # addition
v1 * v2 # multiplication
v1 - v2 # subtraction
sum(v1) # sum over v1
```
Note how the different methods return different outputs. The arithmetic operations return a new vector while `sum` returns a numeric value.

# Vectors
What happens if we try to combine objects of different types using combine?
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
t <- c("a", 1, TRUE)
typeof(t)
t
```

# Vectors
There are lots of commands for generating special types of numeric vectors. For example,
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
N <- 10 # Value to be used in arguments below

seq(N) # generates a sequence from 1 to N
rev(seq(N)) # reverses order
rnorm(N) # samples N times from a normal distribution
rbinom(N,1,0.5) # N observations of a single trial with a 0.5 probability of success
```

# Vectors
We can use the help `?` command to find information about each of these commands.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
?rnorm
```

# Vectors
We can use the `index` to access the specific elements of a vector. R uses square brackets for such indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- rnorm(N)
print(x)
print(x[1]) # R indexing starts at 1; Python and some others start at 0
x[1] <- 9 # We can also use indexing to modify elements
print(x[1])
```

# Vectors
The `head` and `tail` commands are useful when we're working with larger objects.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- rnorm(10000)
length(x)
head(x)
tail(x)
head(x, n=20)
```

# Vectors
Retrieve the final element from `x` using indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }

```


# Vectors
Vectors can also contain `null` elements to indicate missing values, represented by `NA` logical value.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
x <- c(1,2,3,4,NA)
x
is.na(x) # The is.na function indicates whether each value is missing.
```

# Lists
A list is an object that can contain different types of elements, including basic types and vectors.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1.list <- list(v1) # We can easily convert the vector v1 into a list.
```

# Lists
Lists have a more complex form of indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v1.list[1] # The entire vector is considered the first element of the list
v1.list[[1]] # We can access this element by using double brackets
v1.list[[1]][1] # Followed by single brackets to access a specific element
v1.list[1][1] # Otherwise we get the entire vector
```

# Lists
We can easily combine multiple vectors into a list.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v.list <- list(v1,v2) # We could store both vectors in a list
print(v.list)
v.list[[2]][4] # We can use double brackets to get element 4 of list 1
```

# Lists
We can make indexing easier if we start with an empty list and then add elements using a named index with the `$` operator.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
v <- list() # initialize empty list
v$v1 <- v1 # the $ sign allows for named indexing
v$v2 <- v2
print(v)
```

# Lists
Combine `$` and bracket indexing to get the fourth element of `v`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }

```



# Lists
A list could contain a mix of different types. This type of structure forms the backbone of the dataframes we will be using throughout the class.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
cats <- list(c("Gary", "Tabitha"), c(4,1))
print(cats)
```
\tiny See  \href{https://r4ds.had.co.nz/vectors.html}{Chapter 20} of R4DS for more on lists and vectors.

# Matrices
A matrix is a two-dimensional data structure. Like vectors, matrices hold objects of a single type. Here we're defining a matrix using two arguments, the number of rows and columns.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
matrix(nrow=5,ncol=5) # Here there is no content so the matrix is empty
```

# Matrices
A matrix is a two-dimensional data structure.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M <- matrix(0L, nrow=5, ncol=5) # 5x5 matrix of zeros
M
```

# Matrices
We can create a matrix by combining vectors using `cbind` oand `rbind`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
M1 <- cbind(v1,v2) # Treat vectors a columns
print(M1)
M2 <- rbind(v1, v2) # Vectors as rows
print(M2)
```

# Matrices
We can get particular values using two-dimensional indexing.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
dim(M1) # Shows the dimensions of the matrix
i <- 1 # row index
j <- 2 # column index
M1[i,j] # Returns element i,j
M1[i,] # Returns row i
M1[,j] # Returns column i
```

# Matrices
Like lists, we can  also name rows and columns to help make indexing easier. The `colnames` and `rownames` functions show the names of each column and row.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
colnames(M1)
rownames(M1)
```
# Matrices
We can use these functions to assign new names.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
colnames(M1) <- c("X", "Y")
rownames(M1) <- seq(1, nrow(M1))
print(M1)
```

# Data frames
Like its component vectors, a matrix contains data of the same type. If we have a mix of data types we generally want to use a ``data.frame``. Note how the printed version shows the type of each column.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
df <- as.data.frame(M1)
class(df)
df$Z <- c("a","b", "c", "d", "e")
print(df)
```

# Data frames
We can use indexing in the same way as lists to extract elements.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
data(iris) # The `data` function loads a built in dataset
head(iris)
iris$Sepal.Length[1] # Explicitly call column name
iris[[1]][1] # reference column using index
```

# Tibbles
A `tibble` is the `tidyverse` take on a data.frame. It is more "opinionated," which helps to maintain the integrity of your data. It also has some other updated features. We can easily convert any `data.frame` into a `tibble`.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
library(tidyverse) # the library is required to use the as_tibble function
iris.t <- as_tibble(iris) # convert to tibble
class(iris.t)
```
# Tibbles
Tibbles only show the first ten rows when printing (both look the same in RMarkdown, so we have to use the console to compare.)
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize' }
print(iris)
print(iris.t)
```
# Tibbles
Tibbles also tend to provide more warnings when potential issues arise, so they should be less prone to errors than dataframes.
```{r, echo=TRUE, , mysize=TRUE, size='\\footnotesize', warning=T }
iris$year
iris.t$year
```
  
# Questions?


